Modelling writing hesitations in text writing as finite mixture process

### 2023-07-24 report/manuscript.Rmd
Manuscript with appendices separately

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE

@@

header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{threeparttable}
  - \usepackage[normalem]{ulem}
  - \usepackage[utf8]{inputenc}
  - \usepackage{icomma}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
---

```{r packages, include=FALSE}
library(tidyverse)
library(kableExtra)
library(janitor)
library(brms)
library(polspline)
library(patchwork)
library(rmdformats)
library(gridExtra)
library(grid)
library(gtable)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

theme_set(theme_few(base_size = 9) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

beta_label <- "Short key transition\nduration"
delta_label <- "Slowdown for hesitant key\ntransitions"
theta_label <- "Probability of hesitant\ntransitions"

@@

@@

### Uni-modal Gaussian

@@

$$
\begin{align}(\#eq:unimodgaus)
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Uni-modal log-Gaussian

@@

$$
\begin{align}(\#eq:unimodloggaus)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Uni-modal unequal-variance log-Gaussian

$$
\begin{align}(\#eq:unimoduv)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) 
\end{align}
$$

### Bi-modal log-Gaussian (constrained)

@@

$$
\begin{align}(\#eq:bimodcon)
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
\text{where: }
		&\delta \sim \text{N}(0,1)\\
 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$
This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

### Bi-modal log-Gaussian (unconstrained)

@@

$$
\begin{align}(\#eq:bimoduncon)
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: } & \delta \sim \text{N}(0,1)\\
	 	 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta > 0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{align}
$$

@@

Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \@ref(tab:datasets).

```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
                Source = c("@ronneberg2022process",
                              "@torrance2016adolescent",
                              "@torrance",
                              "@rossetti2022s",
                              "@nina_vandermeulen_2020_3893538", 
                           ""),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog",
                           "EyeWrite"),
                `Writing task` = c(
                                 "Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis", 
                                 ""),
                 `N (ppts)` = c(126, 26*2, 39, 47, 658, 45),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                conditions = c(" ", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres",
                               "masked / unmasked"),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch",
                               "Norwegian")) %>% 
  kable(caption = "Datasets in brief.") %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

@@

@@

The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process. The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography.

TODO: might need to remove kids that don't speak Norwegian at home (see github issue).

@@

Data are published in @torrance2016adolescent. Norwegian upper
secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing.

@@

The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group. In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience. Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability. During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e. making it easier to read for a lay audience--by applying what they had learned from their respective training. The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language. It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch.

@@

LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) [@nina_vandermeulen_2020_3893538].

@@

Data are going to be published in @torrance.

Undergraduate university students--*N* = 39, 28 female, mean age = 20.6 years (SD = 1.51)--wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined]. CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste. We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text.

Writing tasks: Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.

@@

@@

The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g. @torrance; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \@ref(tab:keyloc). Keytransitions that terminated in an editing operation were excluded from the analysis. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.

```{r keyloc, results='asis'}
tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
       Description = c("Transitions between any letter",
                       "Keypress after space followed by any letter",
                       "Keypress following a space preceding any letter"),
       Example = c("T\\^h\\^e c\\^a\\^t m\\^e\\^o\\^w\\^e\\^d. T\\^h\\^a\\^t[bsp][bsp]e\\^n i\\^t s\\^l\\^e\\^p\\^t.", 
                   "The \\^cat \\^meowed. That[bsp][bsp]en \\^it \\^slept.", 
                   "The cat meowed. \\^That[bsp][bsp]en it slept.")) %>% 
  kable(caption = "Transition location classification.") %>%
  footnote("'\\^' marks transition location, [bsp] represents backspace. IKIs were timed to the shift keypress.") %>% 
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 12)

@@

```{r reductionfunctions}
source("scripts/get_data_summary.R")
```

```{r}
# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
```

@@

We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \@ref(tab:datareduction).

From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set. This was done for computational reasons to reduce the time the Bayesian models need to complete. For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis. Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant. The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \@ref(tab:datareduction).

@@

extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
          mutate(ds = .x$ds)) %>% 
  mutate(across(where(is.numeric), ~round(.*100, 2)),
         across(starts_with("too"), ~str_c(., "%"))) %>% 
  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
  rename(`$\\le$ 50 msecs` = `<50ms`,
         `$\\ge$ 30 secs` = `>30,000ms`)

random_sample <- 
  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
          mutate(ds = .x$ds)) %>% 
    mutate(across(where(is.numeric), ~round(.*100, 1)),
           across(starts_with("keep"), ~str_c(., "%"))) %>%
    unite("keep", starts_with("keep"), sep = " (") %>% 
    mutate(across(c(keep), ~str_c(., ")"))) %>%
  pivot_wider(names_from = location, values_from = keep)

datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
  relocate(ds) %>% 
  mutate(across(ds, recode, 
                "cato" = "CATO",
                "spl2" = "SPL2", 
                "spl2_shift" = "SPL2 (shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                "gunnexp2" = "GUNNEXP2")) %>% 
  rename(`Dataset` = ds)

kable(datareduction, 
      caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition locattion. Standard error is shown in parentheses.",
        align =c("l", rep("r", 14))) %>%
  kable_styling(position = "center", 
                full_width = T, 
                font_size = 11) %>%
  add_header_above(c(" " = 1, 
                     "Extreme values" = 2,
                     "Randomly sampled data" = 3)) %>%
  column_spec(1, width = "8em")

@@

```{r models, results = 'asis'}
models <- tibble(Models = paste0("M",1:4),
       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
       Equation = paste0("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "}"),
       Description = c("Gaussian distribution",
                       "Log-normal distribution",
                       "Log-normal distribution and unequal variances",
                       "Fluent transition are not allowed to vary by transition location",
                       "Fluent transition are allowed to vary by transition location"))

papaja::apa_table(models, align = c("l", "l", "l"), escape = FALSE, digits = 0, placement = "bp!",
                  caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
                  note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
```

each with random effects for participants. Probability functions used were normal and log-normal in line with typically treatments used in the literature, a log-normal distribution with unequal variances for model predictions, and a bimodal mixed effects model. Stan code for mixture models was based on @roeser2021modelling. Text locations (levels: before sentence, before word, within word) was included as predictor in all models.

@@

For model comparisons we used out-of-sample predictions estimated using Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical]. Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and the difference $\Delta\widehat{elpd}$ between models. The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.

Results for all data sets are shown in Table \@ref(tab:loos). For all data sets we found the same pattern. The mixture of log-normal distributions provided a substantially better fit than uni-modal distribution models. The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across conditions.

@@

```{r}
# Get the differences for tasks / across data set
# use effect size = beta / sd
# averaged across grouping variable

>>>>

documentclass     : "apa7"
classoption       : "man"
output            : 
  papaja::apa6_pdf:
    keep_tex: TRUE
    includes:
      after_body: 
        - "appendix_a.tex"
        - "appendix_b.tex"
        - "appendix_c.tex"
        - "appendix_d.tex"
        - "appendix_e.tex"
        - "appendix_g.tex"
        - "appendix_f.tex"

@@

header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{graphicx}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{threeparttable}
  - \usepackage[normalem]{ulem}
  - \usepackage[utf8]{inputenc}
  - \usepackage{icomma}
  - \usepackage{pdflscape}
  - \newcommand{\blandscape}{\begin{landscape}}
  - \newcommand{\elandscape}{\end{landscape}}
  - \DeclareCaptionFormat{cont}{#1 (cont.)#2#3\par}
---

@@

```{r libs, include=FALSE}
library(tidyverse)
library(janitor)
library(brms)
library(papaja)
library(polspline)
library(patchwork)
library(gridExtra)
library(grid)
library(gtable)
library(scales)
library(ggthemes)
source("../scripts/functions.R")
source("scripts/get_pdens_plot.R")

theme_set(theme_few(base_size = 12) +
            theme(strip.background = element_blank(),
                  legend.position = "top",
                  legend.justification = "right",
                  panel.grid = element_blank(),
                  panel.background = element_rect(fill = "transparent"), # bg of the panel
                  plot.background = element_rect(fill = "transparent", color = NA)))

beta_label <- "Short key transition\nduration (in msecs)"
delta_label <- "Slowdown for hesitant key\ntransitions (in secs)"
theta_label <- "Probability of hesitant\ntransitions"
```

@@

### Unimodal Gaussian

@@

\begin{equation}
\begin{aligned}
(\#eq:unimodgaus)
\text{iki}_i & \sim \text{N}(\mu_i, \sigma_\text{e}^2)\\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2)\\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Unimodal log-Gaussian

@@

\begin{equation}
\begin{aligned}
(\#eq:unimodloggaus)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_\text{e}^2) \\
\text{where: } &
\mu_i = \beta_\text{location[i]} + \text{participant}_i\\
& \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Unimodal unequal-variance log-Gaussian

\begin{equation}
\begin{aligned}
(\#eq:unimoduv)
\text{iki}_i & \sim \text{logN}(\mu_i, \sigma_{e_\text{location[i]}}^2) \\
\text{where: } & \mu_i = \beta_\text{location[i]} + \text{participant}_i\\
 & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
 \text{constraint: } & \sigma_\text{e}^2, \sigma_\text{p}^2>0
\end{aligned}
\end{equation}

### Bimodal log-Gaussian (constrained)

@@

\begin{equation}
\begin{aligned}
(\#eq:bimodcon)
\text{iki}_{i} & \sim \theta_\text{location[i]} \cdot \text{LogN}(\beta + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
\text{where: } & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

```{r}
#		<!-- &\delta \sim \text{N}(0,1)\\ -->

This models takes into account two source = s of participant-specific error: (1) each participant has an individual fluent typing as in the previous models; (2) each participant has in individual hesitation frequency that differs across levels of the categorical predictor.

### Bimodal log-Gaussian (unconstrained)

@@

\begin{equation}
\begin{aligned}
(\#eq:bimoduncon)
\text{iki}_{i} &\sim \theta_\text{location[i]} \cdot \text{LogN}(\beta_\text{location[i]} + \delta_\text{location[i]} + \text{participant}_i, \sigma_{e'_\text{location[i]}}^2) + \\
  & (1 - \theta_\text{location[i]}) \cdot \text{LogN}(\beta_\text{location[i]} + \text{participant}_i, \sigma_{e_\text{location[i]}}^2)\\
	\text{where: }  & \text{participant} \sim \text{N}(0, \sigma_\text{p}^2) \\
\text{constraint: } & \delta, \sigma_{e}^2, \sigma_\text{e'}^2, \sigma_\text{p}^2>0\\
		& \sigma_{e'}^2 > \sigma_{e}^2
\end{aligned}
\end{equation}

```{r}
#<!-- & \delta \sim \text{N}(0,1)\\ -->

@@

Five datasets with keystroke data from text production were used for analysis. An overview can be found in Table \ref{tab:datasets}.

\blandscape
```{r datasets, results='asis'}
tibble(Dataset = c("C2L1", "CATO", "SPL2", "PLanTra", "LIFT", "GUNNEXP2"),
                Source = c("Rønneberg et al. (2022)",
                           "Torrance et al. (2016)",
                           "Torrance et al. (n.d.)",
                           "Rossetti and Van Waes (2022)",
                           "Vandermeulen et al. (2020)", 
                           "Torrance and Ofstad (n.d)"),
                Keylogger = c("EyeWrite", 
                           "EyeWrite", 
                           "CyWrite", 
                           "InputLog", 
                           "InputLog",
                           "EyeWrite"),
                `Writing task` = c("Argumentative essays", 
                                 "Expository texts",
                                 "Argumentative essays",
                                 "Text simplification",
                                 "Synthesis", 
                                 ""),
                 `N (ppts)` = as.character(c(126, 26*2, 39, 47, 658, 45)),
#                n_texts = c(1, 2, 2, 2, NA),
#                n_sentences = c(),
#                n_words = c(),
                `Mean age` = c(11.8, 16.9, 20.6, 23, 16.95, NA),
                 Language = c( "Norwegian", 
                               "Norwegian", 
                               "English (L1) / Spanish (L2)", 
                               "English (L2)", 
                               "Dutch",
                               "Norwegian"),
                Manipulation = c("--", 
                               "weak decoders / control; masked / unmasked",
                               "write in L1 / L2",
                               "pre / post test trained in plain language principles and control", 
                               "Various topics and genres",
                               "masked / unmasked")) %>% 
  arrange(Dataset) %>% 
  apa_table(align = c(rep("l", 4), rep("r", 2), "p{2cm}", "p{3cm}"), 
            escape = FALSE, 
            font_size = "footnotesize",
            caption = "Datasets in brief.")

```
\elandscape

@@

TODO: Mark, can you add information here? I've added a provisional bib entry that needs changing or we just say "unpublished"

The GUNNEXP2 dataset published in @torranceb includes keystroke data from a text composition task performed by Norwegian undergraduate students (mean age: XXX). In this dataset participants wrote text either in a masked condition in which the produced text was replaced by 'x's or in an unmasked condition in which the students composed text as normal. Keystroke data were captured using EyeWrite [@sim07; XXXX].

@@

TODO: Mark, can you add information here?

TODO: might need to remove kids that don't speak Norwegian at home (see github issue)?

The C2L1 data set comprises data Norwegian 6th graders -- *N*=126, mean age 11 years 10 months -- published in @ronneberg2022process. The children composed argumentative essays in Norwegian, a language with a relatively shallow orthography. Keystroke data were captured using EyeWrite [@sim07; XXXX].

@@

TODO: Mark, can you add information here?

Data are published in @torrance2016adolescent. Norwegian upper secondary students--*N*=26, mean age = 16.9 years--with weak decoding skills and 26 age-matched controls composed expository texts by keyboard under two conditions: normally and with letters masked to prevent them reading what they were writing. Keystroke data were captured using EyeWrite [@sim07; XXXX].

@@

TODO: Rianne, can you add information here?

The PLanTra (Plain Language for Financial Content: Assessing the Impact of Training on Students' Revisions and Readers' Comprehension) data set [@rossetti2022s] involved the collection of keystroke data from 47 university students, who were randomly divided into an experimental and a control group. In a pre-test session, all students were assigned an extract of a corporate report dealing with sustainability and were instructed to revise it to make it easier to read for a lay audience. Subsequently, the experimental group received training on how to apply plain language principles to sustainability content, while the control group received training exclusively on the topic of sustainability. During a post-test session, both groups were instructed to revise a second extract of a corporate sustainability report with the same goal--i.e. making it easier to read for a lay audience--by applying what they had learned from their respective training. The texts were in English while the participants were native speakers of other languages (mainly Dutch), so writing took place in second language. It should be pointed out that, while some students decided to revise the assigned texts, the majority of them opted for rewriting the texts from scratch. Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].

@@

TODO: Rianne, can you add information here?

The LIFT (Improving Pre-university Students' Performance in Academic Synthesis Tasks with Level-up Instructions and Feedback Tool) data are published in @vandermeulen2020. Students produced text for 4 topics in two different writing genres (levels: argumentative text, informative text). Keystroke data were captured using InputLog [@leijten2013keystroke; @van2019multilingual; @waes2019].

@@

TODO: Mark, can you add information here?

The SPL2 dataset is published in @torrancea. The data come from a textcomposition task of undergraduate university students -- *N* = 39, 28 female, mean age = 20.6 years (SD = 1.51) -- who wrote two short argumentative essays, one in English (the student's first language in all cases; L1) and one in Spanish (L2) using CyWrite [@chukharev2019combined]. CyWrite provides a writing environment with basic word processing functionality (e.g., Microsoft WordPad), including text selection by mouse action, and copy-and-paste. We recorded the time of each keystroke and mouse action, and tracked writers' eye movements within their emerging text. Participants were given a 40 minute time limit. They wrote essays in response to each of two prompts, with order and L1 / L2 counterbalanced across subjects.

@@

We reanalysed keystroke data from 6 datasets in Bayesian implementations of 5 statistical models. 
Bayesian models, as used in this paper, are ideal for the estimation of parameter values: Bayesian parameter estimates are expresses the uncertainty associated with parameter values as probability distribution [@farrell2018computational; @gelman2014; @lee2014bayesian]. To achieve this, Bayesian models require the explicit inclusion of prior information, i.e. existing knowledge about parameter values, for regulating the parameter space. For small data sets priors influence the inferred parameter estimates (known as posterior); for larger data sets weakly informative and vague priors are quickly overcome by the data [i.e. automatic Ockham's razor, @jefferys1992ockham]. In other words the choice of priors values has less impact on the posterior. In the present paper, we use weakly informative priors to aid model convergence by constraining the parameter space [see e.g. @lambert2018student; @mcelreath2016statistical]. The predictive performance (i.e. fit) of these models is compared in the Results section using leave-one-out cross-validation.

```{r models, results = 'asis'}
models <- tibble(Models = paste0("M",1:5),
       Type = c("LMM", "LMM", "LMM", "MoG", "MoG"),
       `Eq.` = str_c("\ \\ref{eq:", c("unimodgaus", "unimodloggaus", "unimoduv", "bimodcon", "bimoduncon"), "} "),
       Description = c("Gaussian distribution",
                       "Log-normal distribution",
                       "Log-normal distribution and unequal variances",
                       "Fluent transition are not allowed to vary by transition location",
                       "Fluent transition are allowed to vary by transition location"))

apa_table(models, 
          align = c(rep("l", 2), "r", "p{9cm}"), 
          escape = FALSE, 
          digits = 0,
          caption = "Overview of typing-process models. All models were fitted with random intercepts for participants.",
          note = "LMM = Unimodal mixed-effects models; MoG = bimodal mixture of log-Gaussians") 
```

For all models we modelled transition locations (levels: before sentence, before word, within word) was included as predictor in all models. Also, we included the dataset specific experimental conditions as fixed effects and included random intercepts for participants as described in the equations above. We used normal and log-normal probability function in line with typically treatments fouund in the literature. Stan code for mixture models was based on @roeser2021modelling  [see also @vasishth2017; @vasishth2017feature].

@@

The transition types that were analysed in this study focuses on those locations that were found, by previous research, to be psycholinguistically meaningful [e.g. @torrancea; @chukharev2019combined; @torrance2016adolescent; @de2018exploring] and are detailed in Table \ref{tab:keyloc}. Keytransitions that terminated in an editing operation were excluded from the analysis. Transitions that occurred at the beginning of the text or the beginning of a paragraph were not treated as before-sentence transitions.

```{r keyloc, results='asis'}

table <- tibble(`Transition type` = c("Within word", "Below word", "Before sentence"),
       Description = c("Transitions between any letter",
                       "Keypress after space followed by any letter",
                       "Keypress following a space preceding any letter"),
       Example = c("T$^{\\wedge}$h$^{\\wedge}$e c$^{\\wedge}$a$^{\\wedge}$t m$^{\\wedge}$e$^{\\wedge}$o$^{\\wedge}$w$^{\\wedge}$e$^{\\wedge}$d. T$^{\\wedge}$h$^{\\wedge}$a$^{\\wedge}$t[bsp][bsp]e$^{\\wedge}$n i$^{\\wedge}$t s$^{\\wedge}$l$^{\\wedge}$e$^{\\wedge}$p$^{\\wedge}$t.", 
                   "The $^{\\wedge}$cat $^{\\wedge}$meowed. That[bsp][bsp]en $^{\\wedge}$it $^{\\wedge}$slept.", 
                   "The cat meowed. $^{\\wedge}$That[bsp][bsp]en it slept."))

apa_table(table, caption = "Transition location classification.",
            align = c("p{3cm}", "p{3cm}", "p{8cm}"), 
            escape = FALSE, 
            font_size = "footnotesize",
            note = "$'^{\\wedge}$' marks transition location; [bsp] represents backspace. At before-sentence locations, IKIs were timed to the shift keypress.")

@@

```{r reductionfunctions}
source("scripts/get_data_summary.R")

# Get info about data reduction
n_samples <- 100
c2l1 <- get_c2l1("../data/c2l1.csv", n_samples = n_samples)
cato <- get_cato("../data/cato.csv", n_samples = n_samples)
lift <- get_lift("../data/lift.csv", n_samples = 50, n_ppts = 100)
plantra <- get_plantra("../data/plantra.csv", n_samples = n_samples)
spl2 <- get_spl2("../data/spl2.csv", n_samples = n_samples) 
gunnexp2 <- get_gunnexp2("../data/gunnexp2.csv", n_samples = n_samples)
```

@@

We removed keystroke intervals that are extremely short ($\le$ 50 msecs) or extremely long ($\ge$ 30 secs). The percentage of remove keystroke data can be found in Table \ref{tab:datareduction}.

From the remaining data we randomly sampled `r n_samples` observations per participant, per condition, and per transition location, with the exception of the LIFT data set. This was done for computational reasons to reduce the time the Bayesian models need to complete. For the LIFT data set we reduced the number of participants to 100 which is substantially more than most of the other data sets in our analysis. Because we included the large number of writing tasks in the LIFT data set as fixed effect, we sampled 50 observations per condition, location and participant. The percentage of keystroke data that went into the final analysis can be found, by transition location, in Table \ref{tab:datareduction}.

@@

extrem_values <- map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[1]] %>% 
          mutate(ds = .x$ds)) %>% 
  mutate(across(where(is.numeric), ~round(.*100, 2))) %>% 
  unite("<50ms", starts_with("too_fast"), sep = " (") %>% 
  unite(">30,000ms", starts_with("too_slow"), sep = " (") %>% 
  mutate(across(c(`<50ms`, `>30,000ms`), ~str_c(., ")"))) %>% 
  rename(`$\\le$ 50 msecs` = `<50ms`,
         `$\\ge$ 30 secs` = `>30,000ms`)

random_sample <- 
  map_dfr(list(c2l1, cato, lift, plantra, spl2, gunnexp2), ~.x[[2]] %>% 
          mutate(ds = .x$ds)) %>% 
    mutate(across(where(is.numeric), ~round(.*100, 1))) %>%
    unite("keep", starts_with("keep"), sep = " (") %>% 
    mutate(across(c(keep), ~str_c(., ")"))) %>%
  pivot_wider(names_from = location, values_from = keep)

datareduction <- left_join(extrem_values, random_sample, by = "ds") %>% 
  relocate(ds) %>% 
  mutate(across(ds, ~recode(., 
                "cato" = "CATO",
                "spl2" = "SPL2", 
                "spl2_shift" = "SPL2 (shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                "gunnexp2" = "GUNNEXP2"))) %>% 
  rename(`Dataset` = ds)

apa_table(datareduction, 
          escape = FALSE,  
          placement = "bp!",
          caption = "Data reduction. Mean percentage of extreme data removed and the mean percentage of randomly sampled data by transition location. Standard error is shown in parentheses.",
          align =c("l", rep("r", 5)),
          col_spanners = list(" " = 1,
                              "Extreme values in \\%" = c(2, 3),
                              "Randomly sampled data in \\%" = c(4, 6)))

@@

@@

To compare the out-of-sample predictive performance of the models we used Pareto smoothed importance-sampling leave-one-out cross-validation [@vehtari2015pareto; @vehtari2017practical]. Predictive performance was estimated as the sum of the expected log predictive density ($\widehat{elpd}$) and its  difference $\Delta\widehat{elpd}$ between models. The advantage of using leave-one-out cross-validation is that models with more parameters are penalised to prevent overfit.

Results for all data sets are shown in Table \ref{tab:loos}. For all data sets we found the same pattern. The mixture of log-normal distribution models provided a substantially better fit than any of the uni-modal distribution models. The unconstrained version of the mixture of log-normal distributions rendered a higher predictive performance than the constrained version that does not allow the distribution of short keystroke-intervals to vary across transition locations.

We also evaluated to what extent model predictions fit observed data. These comparisons can be found in Appendix \ref{fit-to-data} and echo the findings reported in the model comparisons in Table \ref{tab:loos}.

@@

ps <- purrr::map_dfr(files, ~read_csv(.x) %>% 
  mutate(data = .x)) %>% 
  mutate(across(data, ~sub(".*/([^/]+)/[^/]+\\..*", "\\1", .)),
         across(data, ~recode_factor(., 
                "cato" = "CATO (non-dyslexic\nunmasked)",
                "gunnexp2" = "GUNNEXP2 (unmasked)",
                "spl2" = "SPL2 (L1)", 
#                "spl2 (shift + C)" = "SPL2 (L1; shift + C)", 
                "plantra" = "PLanTra",
                "lift" = "LIFT",
                "c2l1" = "C2L1",
                .ordered = TRUE))) %>% 
  filter(!(str_detect(data, "SPL2") & lang == "ES"),
         !(str_detect(data, "CATO") & group == "dyslexic"),
         !(str_detect(data, "CATO") & task == "masked"),
         !(str_detect(data, "GUNNEXP2") & xn == "masked")) %>% 
  pivot_wider(names_from = param, 
              values_from = value) %>% 
  unnest(cols = beta:theta) %>% 
  mutate(across(c(beta, beta2), ~exp(.)),
         delta = beta2 - beta,
         across(location, ~str_replace(., " ", "\n"))) %>% 
  pivot_longer(beta:theta) %>% 
  summarise(across(value, 
                   list(mean = mean, 
                        lower = lower, 
                        upper = upper),
                   .names = "{.fn}"), 
            .by = c(data, location, name)) %>% 
  filter(name %in% c("beta", "delta", "prob"))
```

As detailed above the mixture model captures the writing process in three conceptually important parameter values: (1) the transition duration associated with fluent writing (which we indicated as $\beta$), (2) the size of the slowdown for hesitant transitions (indicated as $\delta$), and (3) the mixing proportion which represents the probability of hesitant transition durations (indicated as $\theta$).
